---
title: "mETHYLotest EPIC — QC Report"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    self_contained: true
params:
  qc_data: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

```{r check-params}
if (is.null(params$qc_data))
  stop("No QC data provided to the report (qc_data is NULL).")

qc <- params$qc_data

# Helper: section availability badge
avail_badge <- function(obj, label) {
  if (!is.null(obj))
    htmltools::span(
      class = "label label-success",
      style = "margin-right:4px;",
      paste0("\u2713 ", label)
    )
  else
    htmltools::span(
      class = "label label-default",
      style = "margin-right:4px; color:#999;",
      paste0("\u2013 ", label, " (not computed)")
    )
}
```

---

# Dataset Overview

```{r dataset-overview}
# Build summary from first available plot data
n_samples <- if (!is.null(qc$pca_prcomp))
  nrow(qc$pca_prcomp$x) else NA
n_probes  <- if (!is.null(qc$pca_prcomp))
  ncol(qc$pca_prcomp$rotation) else NA

htmltools::div(
  class = "row",
  htmltools::div(
    class = "col-sm-6",
    htmltools::div(
      class = "panel panel-primary",
      htmltools::div(class = "panel-heading",
                     htmltools::strong("QC Sections Computed")),
      htmltools::div(
        class = "panel-body",
        avail_badge(qc$plots$pca_plot,      "PCA"),
        avail_badge(qc$plots$tsne_plot,     "t-SNE"),
        avail_badge(qc$plots$svd_plot,      "SVD"),
        avail_badge(qc$plots$beta_density,  "Beta Distribution"),
        avail_badge(qc$plots$cor_heatmap,   "Sample Correlation"),
        avail_badge(qc$plots$detp_summary,  "Detection P-value"),
        avail_badge(qc$plots$sex_pred,      "Sex Prediction")
      )
    )
  ),
  htmltools::div(
    class = "col-sm-6",
    htmltools::div(
      class = "panel panel-info",
      htmltools::div(class = "panel-heading",
                     htmltools::strong("Report Parameters")),
      htmltools::div(
        class = "panel-body",
        htmltools::tags$table(
          class = "table table-condensed",
          style = "margin:0;",
          htmltools::tags$tr(
            htmltools::tags$td(htmltools::strong("Samples")),
            htmltools::tags$td(if (!is.na(n_samples)) n_samples else "—")
          ),
          htmltools::tags$tr(
            htmltools::tags$td(htmltools::strong("CpG Probes")),
            htmltools::tags$td(if (!is.na(n_probes))
              format(n_probes, big.mark = ",") else "—")
          ),
          htmltools::tags$tr(
            htmltools::tags$td(htmltools::strong("Generated")),
            htmltools::tags$td(format(Sys.time(), "%Y-%m-%d %H:%M:%S"))
          )
        )
      )
    )
  )
)
```

---

# Dimensionality Reduction

## Principal Component Analysis (PCA)

> PC1 and PC2 capture the largest sources of variance across samples.
> Outlier samples appear isolated from the main cluster.

```{r pca-plot, fig.height=6}
if (!is.null(qc$plots$pca_plot)) {
  print(qc$plots$pca_plot)
} else {
  htmltools::div(class = "alert alert-warning",
                 "PCA plot not generated.")
}
```

## t-SNE

> t-SNE is a non-linear dimensionality reduction technique.
> Useful for visualising cluster structure not captured by PCA.
> Note: perplexity was adapted to the sample size.

```{r tsne-plot, fig.height=6}
if (!is.null(qc$plots$tsne_plot)) {
  print(qc$plots$tsne_plot)
} else {
  htmltools::div(class = "alert alert-warning",
                 "t-SNE plot not generated.")
}
```

---

# Covariate Association

## SVD Covariate Correlation Heatmap

> Each cell shows the correlation between a principal component and a
> phenotype variable. Significant associations (BH-corrected) are marked
> with asterisks. Strong associations with technical variables
> (e.g. batch, plate) indicate potential confounding.

```{r svd-plot, fig.height=7}
if (!is.null(qc$plots$svd_plot)) {
  print(qc$plots$svd_plot)
} else {
  htmltools::div(class = "alert alert-warning",
                 "SVD heatmap not generated (requires >= 2 numeric/factor",
                 " covariates beyond reserved columns).")
}
```

---

# Methylation Distribution

## Beta Value Distribution

> Expected shape: bimodal (peaks near 0 = unmethylated, near 1 = methylated).
> A unimodal or flat distribution indicates a problematic sample.
> All curves should overlap — divergent samples are potential outliers.

```{r beta-density, fig.height=6}
if (!is.null(qc$plots$beta_density)) {
  # Native plotly object
  qc$plots$beta_density
} else {
  htmltools::div(class = "alert alert-warning",
                 "Beta density plot not generated.")
}
```

---

# Sample Quality

## Sample Correlation Heatmap

> High-quality cohorts show uniformly high inter-sample correlation
> (typically > 0.95 for EPIC arrays). Samples with systematically
> lower correlation may indicate degraded DNA or labelling errors.
> Samples flagged as potential outliers are noted below the heatmap.

```{r cor-heatmap, fig.height=7}
if (!is.null(qc$plots$cor_heatmap)) {
  # Native plotly object
  qc$plots$cor_heatmap
} else {
  htmltools::div(class = "alert alert-warning",
                 "Correlation heatmap not generated.")
}
```

## Detection P-value Summary

> Probes with detection p-value > threshold are considered unreliable.
> The red dashed line marks the 5% failure threshold — samples exceeding
> this should be considered for removal.

```{r detp-plot, fig.height=5}
if (!is.null(qc$plots$detp_summary)) {
  print(qc$plots$detp_summary)
} else {
  htmltools::div(
    class = "alert alert-warning",
    "Detection p-value summary not generated. ",
    htmltools::tags$br(),
    "This requires ",
    htmltools::tags$code("myLoad$detP"),
    " — use ",
    htmltools::tags$code('method = "minfi"'),
    " in ",
    htmltools::tags$code("champ.load()"),
    " or provide the IDAT directory to compute detP post-hoc."
  )
}
```

## Sex Prediction

> Sex is predicted from median beta values on chrX and chrY probes.
> Mismatches between declared and predicted sex are highlighted in red
> and should be carefully verified.

```{r sex-pred-plot, fig.height=5}
sex_pred <- qc$plots$sex_pred

if (!is.null(sex_pred) && !is.null(sex_pred$plot)) {
  print(sex_pred$plot)
} else {
  htmltools::div(class = "alert alert-warning",
                 "Sex prediction plot not generated.")
}
```

```{r sex-pred-table}
if (!is.null(sex_pred) && !is.null(sex_pred$data)) {

  df_sex <- sex_pred$data

  # Alert for mismatches
  mm <- sex_pred$mismatches
  if (length(mm) > 0L) {
    htmltools::div(
      class = "alert alert-danger",
      htmltools::strong(paste0(length(mm), " sex mismatch(es) detected: ")),
      paste(mm, collapse = ", ")
    )
  } else if ("Match" %in% colnames(df_sex)) {
    htmltools::div(
      class = "alert alert-success",
      "All declared and predicted sex values match."
    )
  }

  # Table
  DT::datatable(
    df_sex,
    rownames = FALSE,
    options  = list(dom = "tp", pageLength = 15, scrollX = TRUE),
    class    = "stripe hover compact"
  ) %>%
    DT::formatStyle(
      if ("Match" %in% colnames(df_sex)) "Match" else character(0),
      backgroundColor = DT::styleEqual(
        c("Match",   "Mismatch"),
        c("#d4edda", "#f8d7da")
      ),
      color = DT::styleEqual(
        c("Match",   "Mismatch"),
        c("#155724", "#721c24")
      ),
      fontWeight = "bold"
    ) %>%
    DT::formatStyle(
      "Predicted_Sex",
      color = DT::styleEqual(
        c("Male",    "Female"),
        c("#1a5276", "#922b21")
      ),
      fontWeight = "bold"
    )

} else {
  htmltools::div(class = "alert alert-warning",
                 "Sex prediction data not available.")
}
```

---

# Session Information

```{r session-info}
si <- sessionInfo()

htmltools::div(
  class = "panel panel-default",
  htmltools::div(class = "panel-heading",
                 htmltools::strong("R Session Info")),
  htmltools::div(
    class = "panel-body",
    htmltools::tags$table(
      class = "table table-condensed table-bordered",
      style = "font-size:12px;",
      htmltools::tags$tbody(
        htmltools::tags$tr(
          htmltools::tags$td(htmltools::strong("R version")),
          htmltools::tags$td(si$R.version$version.string)
        ),
        htmltools::tags$tr(
          htmltools::tags$td(htmltools::strong("Platform")),
          htmltools::tags$td(si$R.version$platform)
        ),
        htmltools::tags$tr(
          htmltools::tags$td(htmltools::strong("Key packages")),
          htmltools::tags$td(
            paste(
              sapply(
                intersect(c("mETHYLotest", "ChAMP", "minfi", "PCAtools",
                            "Rtsne", "plotly", "shiny"),
                          names(si$otherPkgs)),
                function(p) paste0(p, " v", si$otherPkgs[[p]]$Version)
              ),
              collapse = ", "
            )
          )
        )
      )
    )
  )
)
```
