---
title: "mETHYLotest EPIC Report"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: { collapsed: false }
    toc_depth: 3
    number_sections: true
    code_folding: hide
    fig_width: 10
    fig_height: 7
    self_contained: true
params:
  project_dir: ""
  beta: !r NULL
  pheno: !r NULL
  cfg: !r NULL
  dmp_list: !r NULL
  signatures_dir: ""
  validation_results: !r NULL
  top_n: 50
---

<style>
h1.title{color:#8e44ad}
h1{border-bottom:3px solid #8e44ad;padding-bottom:8px}
h2{color:#2c3e50}

.stat-box{display:inline-block;text-align:center;padding:15px 30px;
  border-radius:8px;color:#fff;margin:5px;min-width:130px}
.stat-box .big-num{margin:0;font-size:32px;font-weight:bold;line-height:1.2}
.stat-box .label{margin:0;font-size:12px;opacity:.85}

.sig-card{display:inline-block;text-align:center;padding:12px 20px;
  border-radius:8px;color:#fff;margin:5px;min-width:120px}
.sig-card .big-num{margin:0;font-size:24px;font-weight:bold;line-height:1.2}
.sig-card .label{margin:0;font-size:11px;opacity:.85}

.info-box{background:#eaf2f8;border-left:4px solid #3498db;
  padding:12px 15px;border-radius:4px;margin:10px 0}
.warn-box{background:#fdf2e9;border-left:4px solid #e67e22;
  padding:12px 15px;border-radius:4px;margin:10px 0}

img{max-width:100%!important;height:auto!important}
.figure{max-width:100%;overflow-x:auto}
.plotly,.html-widget{max-width:100%!important;overflow-x:auto}
pre{overflow-x:auto}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%", dpi = 150
)

suppressPackageStartupMessages({ library(ggplot2) })

has_plotly    <- requireNamespace("plotly",      quietly = TRUE)
has_rtsne    <- requireNamespace("Rtsne",       quietly = TRUE)
has_mstats   <- requireNamespace("matrixStats", quietly = TRUE)
has_patchwork<- requireNamespace("patchwork",   quietly = TRUE)
has_proc     <- requireNamespace("pROC",        quietly = TRUE)
has_scales   <- requireNamespace("scales",      quietly = TRUE)

beta     <- params$beta
pheno    <- params$pheno
cfg      <- params$cfg
dmp_list <- params$dmp_list
top_n    <- params$top_n

if (is.null(beta))    stop("beta is NULL")
if (!is.matrix(beta)) stop("beta is not a matrix, it is: ", class(beta)[1])
if (is.null(pheno))   stop("pheno is NULL")
if (is.null(cfg))     stop("cfg is NULL")

has_dmp <- is.list(dmp_list) && length(dmp_list) > 0

# ═══════════════════════════════════════════════════════
# Build groups vector
# ═══════════════════════════════════════════════════════
grp_var <- "Sample_Group"
if (!is.null(cfg$analysis_pheno) && nzchar(cfg$analysis_pheno)) {
  grp_var <- cfg$analysis_pheno
}

if (is.data.frame(pheno)) {
  if (!"Sample_Name" %in% colnames(pheno)) {
    if (!is.null(rownames(pheno)) &&
        !identical(rownames(pheno), as.character(seq_len(nrow(pheno))))) {
      pheno$Sample_Name <- rownames(pheno)
    } else {
      pheno$Sample_Name <- colnames(beta)[seq_len(nrow(pheno))]
    }
  }
  if (!grp_var %in% colnames(pheno) && "Sample_Group" %in% colnames(pheno)) {
    grp_var <- "Sample_Group"
  }
  if (grp_var %in% colnames(pheno)) {
    pheno_vec <- setNames(as.character(pheno[[grp_var]]),
                          as.character(pheno$Sample_Name))
  } else {
    other_cols <- setdiff(colnames(pheno), "Sample_Name")
    if (length(other_cols) > 0) {
      pheno_vec <- setNames(as.character(pheno[[other_cols[1]]]),
                            as.character(pheno$Sample_Name))
    } else {
      pheno_vec <- setNames(rep("Unknown", nrow(pheno)),
                            as.character(pheno$Sample_Name))
    }
  }
} else if (is.character(pheno) || is.factor(pheno)) {
  if (!is.null(names(pheno))) {
    pheno_vec <- setNames(as.character(pheno), names(pheno))
  } else {
    pheno_vec <- setNames(as.character(pheno), colnames(beta))
  }
} else {
  pheno_vec <- setNames(rep("Unknown", ncol(beta)), colnames(beta))
}

for (s in setdiff(colnames(beta), names(pheno_vec))) {
  orig <- sub("_copy[0-9]*$", "", s)
  if (orig %in% names(pheno_vec)) {
    pheno_vec[s] <- pheno_vec[orig]
  } else {
    pheno_vec[s] <- "Unknown"
  }
}

pheno_vec <- pheno_vec[colnames(beta)]
pheno_vec[is.na(pheno_vec)] <- "Unknown"
groups <- unname(pheno_vec)
glvl   <- unique(groups)
ng     <- length(glvl)

# ═══════════════════════════════════════════════════════
# Palette
# ═══════════════════════════════════════════════════════
base_cols <- c("#e74c3c","#3498db","#2ecc71","#f39c12",
               "#9b59b6","#1abc9c","#e67e22","#34495e")
if (ng <= length(base_cols)) {
  pal <- setNames(base_cols[seq_len(ng)], glvl)
} else if (has_scales) {
  pal <- setNames(scales::hue_pal()(ng), glvl)
} else {
  pal <- setNames(rainbow(ng), glvl)
}

adj_thr <- if (!is.null(cfg$dmp_adj_p_val)) cfg$dmp_adj_p_val else 0.05
db_thr  <- 0.1

# ═══════════════════════════════════════════════════════
# Helpers
# ═══════════════════════════════════════════════════════
row_vars <- function(m) {
  if (has_mstats) {
    matrixStats::rowVars(m, na.rm = TRUE)
  } else {
    apply(m, 1, var, na.rm = TRUE)
  }
}

top_var_idx <- function(m, n = 10000) {
  v <- row_vars(m)
  v[!is.finite(v)] <- 0
  head(order(v, decreasing = TRUE), min(n, nrow(m)))
}

impute_row_mean <- function(m) {
  if (!any(is.na(m))) return(m)
  rm <- rowMeans(m, na.rm = TRUE)
  rm[is.na(rm)] <- 0.5
  idx <- which(is.na(m), arr.ind = TRUE)
  if (nrow(idx) > 0) {
    m[idx] <- rm[idx[, 1]]
  }
  return(m)
}

lfc_col <- function(d) {
  for (cn in c("logFC", "deltaBeta", "delta_beta")) {
    if (cn %in% colnames(d)) return(cn)
  }
  return(NULL)
}

val_field <- function(v, fields, default = NULL) {
  for (f in fields) {
    if (!is.null(v[[f]])) return(v[[f]])
  }
  return(default)
}

# ═══════════════════════════════════════════════════════
# Render helpers
# ═══════════════════════════════════════════════════════
.plot_counter <- 0L

render_plot <- function(p, fig.width = 10, fig.height = 7) {
  .plot_counter <<- .plot_counter + 1L
  p_env <- new.env(parent = globalenv())
  p_env$p <- p
  chunk_id <- paste0("auto_plot_", .plot_counter)

  if (inherits(p, "gtable") || inherits(p, "grob")) {
    ct <- c(
      sprintf("```{r %s, echo=FALSE, warning=FALSE, message=FALSE, fig.width=%s, fig.height=%s, out.width='100%%'}",
              chunk_id, fig.width, fig.height),
      "grid::grid.newpage()",
      "grid::grid.draw(p)",
      "```")
  } else {
    ct <- c(
      sprintf("```{r %s, echo=FALSE, warning=FALSE, message=FALSE, fig.width=%s, fig.height=%s, out.width='100%%'}",
              chunk_id, fig.width, fig.height),
      "print(p)",
      "```")
  }
  cat(knitr::knit_child(text = ct, envir = p_env, quiet = TRUE))
}

stat_box <- function(value, label, bg) {
  sprintf('<div class="stat-box" style="background:%s;"><span class="big-num">%s</span><br><span class="label">%s</span></div>',
          bg, value, label)
}

sig_card <- function(value, label, bg) {
  sprintf('<div class="sig-card" style="background:%s;"><span class="big-num">%s</span><br><span class="label">%s</span></div>',
          bg, value, label)
}

# ═══════════════════════════════════════════════════════
# Validation helpers
# ═══════════════════════════════════════════════════════
find_val_result <- function(sig_name, val_res) {
  if (!is.list(val_res) || length(val_res) == 0) return(NULL)
  vn <- names(val_res)
  if (sig_name %in% vn) return(val_res[[sig_name]])
  nm_txt <- paste0(sig_name, ".txt")
  if (nm_txt %in% vn) return(val_res[[nm_txt]])
  nm_strip <- gsub("^DMP_default_", "", sig_name)
  if (nm_strip %in% vn) return(val_res[[nm_strip]])
  nm_strip_txt <- paste0(nm_strip, ".txt")
  if (nm_strip_txt %in% vn) return(val_res[[nm_strip_txt]])
  idx <- which(tolower(vn) == tolower(sig_name))
  if (length(idx) > 0) return(val_res[[idx[1]]])
  idx <- which(tolower(vn) == tolower(nm_strip))
  if (length(idx) > 0) return(val_res[[idx[1]]])
  idx <- grep(sig_name, vn, fixed = TRUE)
  if (length(idx) > 0) return(val_res[[idx[1]]])
  idx <- grep(nm_strip, vn, fixed = TRUE)
  if (length(idx) > 0) return(val_res[[idx[1]]])
  return(NULL)
}

extract_metrics <- function(v) {
  m <- list()
  for (nm in c("mean_accuracy","accuracy","overall_accuracy")) {
    if (!is.null(v[[nm]]) && is.numeric(v[[nm]]) && length(v[[nm]])==1) { m$accuracy <- v[[nm]]; break }
  }
  for (nm in c("sd_accuracy","sd_acc")) {
    if (!is.null(v[[nm]]) && is.numeric(v[[nm]])) { m$sd_accuracy <- v[[nm]]; break }
  }
  for (nm in c("auc","AUC","mean_auc")) {
    if (!is.null(v[[nm]]) && is.numeric(v[[nm]])) { m$auc <- v[[nm]]; break }
  }
  for (nm in c("sensitivity","sens","Sensitivity","mean_sensitivity")) {
    if (!is.null(v[[nm]]) && is.numeric(v[[nm]])) { m$sensitivity <- v[[nm]]; break }
  }
  for (nm in c("specificity","spec","Specificity","mean_specificity")) {
    if (!is.null(v[[nm]]) && is.numeric(v[[nm]])) { m$specificity <- v[[nm]]; break }
  }
  sil <- v$silhouette
  if (!is.null(sil)) {
    if (is.numeric(sil) && length(sil)==1) {
      m$silhouette <- sil
    } else if (is.list(sil) && !is.null(sil$avg.width)) {
      m$silhouette <- sil$avg.width
    } else if (is.matrix(sil) || inherits(sil,"silhouette")) {
      ss <- tryCatch(summary(sil), error=function(e) NULL)
      if (!is.null(ss) && !is.null(ss$avg.width)) {
        m$silhouette <- ss$avg.width
        m$silhouette_per_cluster <- ss$clus.avg.widths
      } else if (is.matrix(sil) && ncol(sil)>=3) {
        m$silhouette <- mean(sil[,3], na.rm=TRUE)
      }
    }
  }
  svm <- v$svm_model
  if (!is.null(svm)) {
    m$has_svm <- TRUE
    if (!is.null(svm$tot.accuracy)) m$svm_accuracy <- svm$tot.accuracy/100
    if (!is.null(svm$accuracies)) {
      m$svm_cv_accuracies <- svm$accuracies/100
      if (is.null(m$accuracy)) m$accuracy <- mean(svm$accuracies/100)
    }
    m$svm_kernel <- if (!is.null(svm$kernel)) as.character(svm$kernel) else "unknown"
    m$svm_nSV <- svm$nSV
  }
  pca_obj <- v$pca
  if (!is.null(pca_obj)) {
    m$has_pca <- TRUE
    if (inherits(pca_obj,"prcomp")) {
      ve <- pca_obj$sdev^2; m$pca_var_pct <- round(100*ve/sum(ve),1); m$pca_obj <- pca_obj
    } else if (is.list(pca_obj) && !is.null(pca_obj$sdev)) {
      ve <- pca_obj$sdev^2; m$pca_var_pct <- round(100*ve/sum(ve),1); m$pca_obj <- pca_obj
    }
  }
  for (nm in c("fold_accuracies","cv_accuracy","accuracy_per_fold","fold_acc")) {
    if (!is.null(v[[nm]]) && is.numeric(v[[nm]]) && length(v[[nm]])>1) { m$fold_accuracies <- v[[nm]]; break }
  }
  for (nm in c("confusion","confusion_matrix","conf_matrix")) {
    if (!is.null(v[[nm]])) { m$confusion <- v[[nm]]; break }
  }
  for (nm in c("roc","roc_obj","ROC")) {
    if (!is.null(v[[nm]])) { m$roc <- v[[nm]]; break }
  }
  if (!is.null(v$metric_used)) m$metric_used <- v$metric_used
  if (!is.null(v$n_cpgs)) m$n_cpgs <- v$n_cpgs
  return(m)
}

# ═══════════════════════════════════════════════════════
# Prepare DMP data.frame
# ═══════════════════════════════════════════════════════
prepare_dmp_df <- function(d) {
  if ("cgi" %in% colnames(d)) colnames(d)[colnames(d)=="cgi"] <- "CGI"
  if (!("CGI" %in% colnames(d))) d$CGI <- "N/A"
  if (!("feature" %in% colnames(d))) d$feature <- "N/A"
  if (!("P.Value" %in% colnames(d))) {
    if ("P.Val" %in% colnames(d)) { d$P.Value <- d$P.Val }
    else if ("adj.P.Val" %in% colnames(d)) { d$P.Value <- d$adj.P.Val }
    else { d$P.Value <- 1 }
  }
  lc <- lfc_col(d)
  if (!is.null(lc)) { d$deltaBeta <- d[[lc]] } else { d$deltaBeta <- 0 }
  d$deltaBeta[is.na(d$deltaBeta)] <- 0
  if (!("adj.P.Val" %in% colnames(d))) d$adj.P.Val <- 1
  d$adj.P.Val[is.na(d$adj.P.Val)] <- 1
  d$Status <- ifelse(d$adj.P.Val < adj_thr & d$deltaBeta > db_thr, "Hyper",
              ifelse(d$adj.P.Val < adj_thr & d$deltaBeta < -db_thr, "Hypo", "NS"))
  return(d)
}

# ═══════════════════════════════════════════════════════
# Render DMP panels (volcano, density, distributions,
# feature x CGI, top DMPs)
# heading: "###" or "####"
# ═══════════════════════════════════════════════════════
render_dmp_panels <- function(df, label, heading = "###") {

  n_h <- sum(df$Status=="Hyper"); n_ho <- sum(df$Status=="Hypo")
  df_sig <- df[df$Status != "NS", ]

  # Volcano
  cat(sprintf("\n%s Volcano\n\n", heading))
  df_ns <- df[df$Status=="NS", ]
  if (nrow(df_ns)>5000) df_ns <- df_ns[sample(nrow(df_ns),5000), ]
  df_viz <- rbind(df_sig, df_ns)
  df_viz$Status <- factor(df_viz$Status, levels=c("Hyper","Hypo","NS"))
  p <- ggplot(df_viz, aes(deltaBeta, -log10(adj.P.Val), colour=Status)) +
    geom_point(alpha=.5, size=.8) +
    scale_colour_manual(values=c(Hyper="#d9534f",Hypo="#0275d8",NS="grey70"), drop=FALSE) +
    geom_hline(yintercept=-log10(adj_thr), linetype="dashed", colour="grey40") +
    geom_vline(xintercept=c(-db_thr, db_thr), linetype="dotted", colour="grey60") +
    labs(title=sprintf("%s - %d Hyper / %d Hypo", label, n_h, n_ho),
         x="Delta Beta", y="-log10(adj. P)") +
    theme_minimal(base_size=12)
  render_plot(p, fig.width=10, fig.height=5)
  cat("\n\n")

  # Density
  cat(sprintf("\n%s Density\n\n", heading))
  if (nrow(df_sig)>0) {
    p <- ggplot(df_sig, aes(deltaBeta, fill=Status)) + geom_density(alpha=.6) +
      scale_fill_manual(values=c(Hyper="#d9534f",Hypo="#0275d8")) +
      labs(title=sprintf("Delta Beta Density - %s", label), x="Delta Beta", y="Density") +
      theme_minimal(base_size=12)
    render_plot(p, fig.width=10, fig.height=5)
  } else { cat("No significant DMPs.\n") }
  cat("\n\n")

  # Distributions
  cat(sprintf("\n%s Distributions\n\n", heading))
  if (nrow(df_sig)>0) {
    pp <- list()
    tbl_cgi <- as.data.frame(table(df_sig$CGI, df_sig$Status), stringsAsFactors=FALSE)
    colnames(tbl_cgi) <- c("CGI","Status","N"); tbl_cgi <- tbl_cgi[tbl_cgi$N>0,]
    if (nrow(tbl_cgi)>0)
      pp[[length(pp)+1]] <- ggplot(tbl_cgi, aes(reorder(CGI,N), N, fill=Status)) +
        geom_col(position="dodge") + scale_fill_manual(values=c(Hyper="#d9534f",Hypo="#0275d8")) +
        coord_flip() + labs(title="CGI", x=NULL, y="Count") +
        theme_minimal(base_size=10) + theme(legend.position="none")

    tbl_feat <- as.data.frame(table(df_sig$feature, df_sig$Status), stringsAsFactors=FALSE)
    colnames(tbl_feat) <- c("Feature","Status","N"); tbl_feat <- tbl_feat[tbl_feat$N>0,]
    if (nrow(tbl_feat)>0)
      pp[[length(pp)+1]] <- ggplot(tbl_feat, aes(reorder(Feature,N), N, fill=Status)) +
        geom_col(position="dodge") + scale_fill_manual(values=c(Hyper="#d9534f",Hypo="#0275d8")) +
        coord_flip() + labs(title="Gene Feature", x=NULL, y="Count") +
        theme_minimal(base_size=10) + theme(legend.position="none")

    if ("CHR" %in% colnames(df_sig)) {
      chr_f <- factor(gsub("chr","", as.character(df_sig$CHR), ignore.case=TRUE),
                      levels=c(as.character(1:22),"X","Y"))
      tbl_chr <- as.data.frame(table(chr_f, df_sig$Status), stringsAsFactors=FALSE)
      colnames(tbl_chr) <- c("Chr","Status","N"); tbl_chr <- tbl_chr[tbl_chr$N>0,]
      if (nrow(tbl_chr)>0)
        pp[[length(pp)+1]] <- ggplot(tbl_chr, aes(Chr, as.integer(N), fill=Status)) +
          geom_col(position="dodge") + scale_fill_manual(values=c(Hyper="#d9534f",Hypo="#0275d8")) +
          labs(title="Chromosome", x=NULL, y="Count") +
          theme_minimal(base_size=10) +
          theme(axis.text.x=element_text(angle=45,hjust=1), legend.position="bottom")
    }
    if (length(pp)>0) {
      if (has_patchwork) {
        render_plot(patchwork::wrap_plots(pp, ncol=min(3,length(pp))), fig.width=12, fig.height=5)
      } else {
        for (plt in pp) render_plot(plt, fig.width=6, fig.height=5)
      }
    }
  } else { cat("No significant DMPs.\n") }
  cat("\n\n")

  # Feature x CGI
  cat(sprintf("\n%s Feature x CGI\n\n", heading))
  if (nrow(df_sig)>0 && any(df_sig$CGI!="N/A") && any(df_sig$feature!="N/A")) {
    p <- ggplot(df_sig, aes(feature, fill=CGI)) +
      geom_bar(position="fill") + facet_wrap(~Status) +
      scale_fill_viridis_d() + labs(title=label, y="Proportion") +
      theme_minimal(base_size=11) + theme(axis.text.x=element_text(angle=45,hjust=1))
    render_plot(p, fig.width=10, fig.height=5)
  } else { cat("No annotation data.\n") }
  cat("\n\n")

  # Top DMPs
  cat(sprintf("\n%s Top DMPs\n\n", heading))
  df_top <- df_sig[order(df_sig$adj.P.Val), ]
  df_top <- head(df_top, top_n)
  if (nrow(df_top)>0) {
    df_top$CpG <- rownames(df_top)
    show_cols <- intersect(
      c("CpG","deltaBeta","adj.P.Val","P.Value","CHR","MAPINFO",
        "gene","feature","CGI","Status",
        "UCSC_RefGene_Name","UCSC_RefGene_Group","Relation_to_Island"),
      colnames(df_top))
    for (pc in intersect(c("P.Value","adj.P.Val"), show_cols)) {
      df_top[[pc]] <- signif(df_top[[pc]], 3)
    }
    print(knitr::kable(df_top[, show_cols, drop=FALSE], format="html", row.names=FALSE,
      table.attr='class="table table-striped table-condensed" style="font-size:12px;"'))
  } else { cat("No significant DMPs.\n") }
  cat("\n\n")
}

# ═══════════════════════════════════════════════════════
# Load signatures
# ═══════════════════════════════════════════════════════
signatures  <- list()
sig_sources <- character(0)
has_sigs    <- FALSE

sig_dir_param <- params$signatures_dir
if (!is.null(sig_dir_param) && nzchar(sig_dir_param) && dir.exists(sig_dir_param)) {
  sig_files <- list.files(sig_dir_param, pattern="\\.txt$", full.names=TRUE)
  for (f in sig_files) {
    nm <- tools::file_path_sans_ext(basename(f))
    cpgs <- readLines(f, warn=FALSE)
    cpgs <- cpgs[nzchar(trimws(cpgs))]
    if (length(cpgs)>0) {
      signatures[[nm]] <- cpgs
      sig_sources[nm] <- if (grepl("^DMP_default_", nm)) "Default" else "User"
    }
  }
  has_sigs <- length(signatures) > 0
}

val_res <- params$validation_results
has_val <- is.list(val_res) && length(val_res) > 0
```

<!-- ================================================================== -->
<!-- 1. PROJECT OVERVIEW                                                 -->
<!-- ================================================================== -->

# Project Overview

```{r overview, results='asis'}
cat('<div style="display:flex;flex-wrap:wrap;gap:8px;margin:15px 0;">')
cat(stat_box(ncol(beta), "Samples", "#8e44ad"))
cat(stat_box(format(nrow(beta), big.mark=","), "CpG Probes", "#3498db"))
cat(stat_box(ng, "Groups", "#27ae60"))
if (has_dmp) {
  ns <- sum(sapply(dmp_list, function(d) {
    if ("adj.P.Val" %in% colnames(d)) sum(d$adj.P.Val < adj_thr, na.rm=TRUE) else 0L
  }))
  cat(stat_box(format(ns, big.mark=","), "Sig. DMPs", "#e74c3c"))
}
if (has_sigs) cat(stat_box(length(signatures), "Signatures", "#e67e22"))
cat('</div>')
```

## Configuration {-}

```{r config_tbl}
safe <- function(x, fb="---") {
  if (is.null(x)||length(x)==0||all(is.na(x))) return(fb)
  paste(as.character(x), collapse=", ")
}
cfg_df <- data.frame(
  Parameter = c("Project","Normalisation","Imputation","Cell-type",
                "Batch","DMP","DMR","Block","GSEA","CNA"),
  Value = c(
    safe(cfg$project_name), safe(cfg$norm_method),
    if (isTRUE(cfg$do_impute)) safe(cfg$impute_method) else "No",
    if (isTRUE(cfg$do_refbase)) "Yes (blood)" else "No",
    if (isTRUE(cfg$perform_batch_correction)) safe(cfg$batch_cols) else "No",
    if (isTRUE(cfg$do_dmp)) paste0("adj.P < ", safe(cfg$dmp_adj_p_val,"0.05")) else "No",
    if (isTRUE(cfg$do_dmr)) safe(cfg$dmr_method) else "No",
    if (isTRUE(cfg$do_block)) "Yes" else "No",
    if (isTRUE(cfg$do_gsea)) safe(cfg$gsea_method) else "No",
    if (isTRUE(cfg$do_cna)) "Yes" else "No"),
  stringsAsFactors=FALSE)
knitr::kable(cfg_df, format="html", table.attr='class="table table-striped table-condensed"')
```

## Samples per Group {-}

```{r grp_tbl}
grp_tbl <- as.data.frame(table(Group=groups), stringsAsFactors=FALSE)
colnames(grp_tbl) <- c("Group","N")
knitr::kable(grp_tbl, format="html", table.attr='class="table table-bordered"')
```

<!-- ================================================================== -->
<!-- 2. QUALITY CONTROL                                                  -->
<!-- ================================================================== -->

# Quality Control {.tabset}

## Beta Density

```{r beta_density, fig.height=5}
set.seed(42)
n_probes <- min(50000, nrow(beta))
idx <- sample(nrow(beta), n_probes)
bs <- beta[idx, , drop=FALSE]

df_d <- do.call(rbind, lapply(seq_len(ncol(bs)), function(j) {
  vals <- bs[,j]; vals <- vals[is.finite(vals)]
  if (length(vals)==0) return(NULL)
  data.frame(b=vals, Sample=colnames(bs)[j], Group=groups[j], stringsAsFactors=FALSE)
}))

if (!is.null(df_d) && nrow(df_d)>0) {
  p <- ggplot(df_d, aes(b, colour=Group, group=Sample)) +
    geom_density(linewidth=.35, alpha=.4) + scale_colour_manual(values=pal) +
    labs(title="Beta-value Density", x=expression(beta), y="Density") +
    theme_minimal(base_size=13) + theme(legend.position="bottom")
  if (has_plotly) plotly::ggplotly(p) else p
} else { cat("No valid beta values.\n") }
```

## PCA {.tabset}

```{r pca_compute}
idx_v <- top_var_idx(beta, 10000)
bpca <- impute_row_mean(beta[idx_v,,drop=FALSE])
pca <- prcomp(t(bpca), scale.=TRUE, center=TRUE)
vpct <- round(100*pca$sdev^2/sum(pca$sdev^2), 1)
n_pc <- min(5, ncol(pca$x))
pcdf <- data.frame(pca$x[,seq_len(n_pc),drop=FALSE], Group=groups, Sample=colnames(beta))
```

### PC1 vs PC2

```{r pca12, fig.height=6}
p <- ggplot(pcdf, aes(PC1, PC2, colour=Group, label=Sample)) +
  geom_point(size=3, alpha=.8) + scale_colour_manual(values=pal) +
  labs(title="PCA - PC1 vs PC2", x=sprintf("PC1 (%s%%)", vpct[1]),
       y=sprintf("PC2 (%s%%)", vpct[2])) +
  theme_minimal(base_size=13) + theme(legend.position="bottom")
grp_counts <- table(groups)
if (any(grp_counts>=3))
  p <- p + stat_ellipse(level=.90, linetype="dashed", linewidth=.6, show.legend=FALSE,
                         data=pcdf[groups %in% names(grp_counts[grp_counts>=3]),])
if (has_plotly) plotly::ggplotly(p) else p
```

### PC2 vs PC3

```{r pca23, fig.height=6}
if (n_pc>=3) {
  p <- ggplot(pcdf, aes(PC2, PC3, colour=Group, label=Sample)) +
    geom_point(size=3, alpha=.8) + scale_colour_manual(values=pal) +
    labs(title="PCA - PC2 vs PC3", x=sprintf("PC2 (%s%%)", vpct[2]),
         y=sprintf("PC3 (%s%%)", vpct[3])) +
    theme_minimal(base_size=13) + theme(legend.position="bottom")
  if (has_plotly) plotly::ggplotly(p) else p
} else { cat("Fewer than 3 PCs.") }
```

### Scree Plot

```{r scree, fig.height=4}
ns <- min(15, length(vpct))
sdf <- data.frame(PC=seq_len(ns), V=vpct[seq_len(ns)])
ggplot(sdf, aes(PC, V)) + geom_col(fill="#8e44ad", alpha=.8) +
  geom_line(colour="#e74c3c", linewidth=1) + geom_point(colour="#e74c3c", size=2) +
  labs(title="Scree Plot", x="PC", y="Variance (%)") +
  scale_x_continuous(breaks=sdf$PC) + theme_minimal(base_size=13)
```

## t-SNE

```{r tsne, fig.height=6}
tsne_out <- NULL
if (has_rtsne && ncol(beta)>=4) {
  bt <- t(impute_row_mean(beta[top_var_idx(beta,5000),,drop=FALSE]))
  dups <- duplicated(bt); n_unique <- sum(!dups)
  perp <- min(30, floor((n_unique-1)/3))
  if (perp>=1 && n_unique>=4) {
    set.seed(42)
    tsne <- Rtsne::Rtsne(bt[!dups,,drop=FALSE], perplexity=perp,
                          check_duplicates=FALSE, verbose=FALSE)
    tdf <- data.frame(D1=tsne$Y[,1], D2=tsne$Y[,2],
                       Group=groups[!dups], Sample=colnames(beta)[!dups])
    tsne_out <- ggplot(tdf, aes(D1, D2, colour=Group, label=Sample)) +
      geom_point(size=3, alpha=.8) + scale_colour_manual(values=pal) +
      labs(title=sprintf("t-SNE (perplexity=%d)", perp), x="t-SNE 1", y="t-SNE 2") +
      theme_minimal(base_size=13) + theme(legend.position="bottom")
  } else { cat("Not enough unique samples.") }
} else { cat("Rtsne not installed or < 4 samples.") }
if (!is.null(tsne_out)) { if (has_plotly) plotly::ggplotly(tsne_out) else tsne_out }
```

<!-- ================================================================== -->
<!-- 3. DMP ANALYSIS                                                     -->
<!-- ================================================================== -->

```{r dmp_gate, results='asis'}
if (!has_dmp) {
  cat("\n# DMP Analysis\n\n",'<div class="info-box">No DMP results provided.</div>\n')
}
```

```{r dmp_content, eval=has_dmp, results='asis', fig.width=10, fig.height=5}
cat("\n# DMP Analysis {.tabset}\n\n")

# ══════════════════════════════════════════════════════
# We merge all comparisons into one df_full
# (for most projects there is a single comparison)
# ══════════════════════════════════════════════════════
all_dmp_dfs <- lapply(names(dmp_list), function(nm) {
  d <- prepare_dmp_df(dmp_list[[nm]])
  d$.comparison <- nm
  return(d)
})
df_full <- do.call(rbind, all_dmp_dfs)

n_comparisons <- length(dmp_list)
sig_cpg_names <- rownames(df_full)[df_full$Status != "NS"]

# ══════════════════════════════════════════════════════
# ## Summary
# ══════════════════════════════════════════════════════
cat("## Summary\n\n")

# Stat cards
n_total <- nrow(df_full)
n_sig   <- sum(df_full$Status != "NS")
n_hyper <- sum(df_full$Status == "Hyper")
n_hypo  <- sum(df_full$Status == "Hypo")

cat('<div style="display:flex;flex-wrap:wrap;gap:8px;margin:10px 0;">\n')
cat(stat_box(format(n_total, big.mark=","), "Total CpGs Tested", "#8e44ad"))
cat(stat_box(format(n_sig, big.mark=","), "Significant", "#e74c3c"))
cat(stat_box(format(n_hyper, big.mark=","), "Hyper", "#d9534f"))
cat(stat_box(format(n_hypo, big.mark=","), "Hypo", "#0275d8"))
if (n_comparisons > 1) {
  cat(stat_box(n_comparisons, "Comparisons", "#27ae60"))
}
if (has_sigs) {
  cat(stat_box(length(signatures), "Signatures", "#e67e22"))
}
cat('</div>\n\n')

# Per-comparison table
if (n_comparisons >= 1) {
  summ <- do.call(rbind, lapply(names(dmp_list), function(nm) {
    d <- df_full[df_full$.comparison == nm, ]
    data.frame(Comparison = nm,
               Total      = nrow(d),
               Significant= sum(d$Status != "NS"),
               Hyper      = sum(d$Status == "Hyper"),
               Hypo       = sum(d$Status == "Hypo"),
               stringsAsFactors = FALSE)
  }))
  print(knitr::kable(summ, format="html",
    table.attr='class="table table-bordered table-striped"'))
  cat("\n\n")
}

# Signatures overlap table
if (has_sigs && length(sig_cpg_names) > 0) {
  cat("**Signatures overlap:**\n\n")
  ovl_df <- do.call(rbind, lapply(names(signatures), function(sn) {
    sc <- signatures[[sn]]
    in_dmp <- sum(sc %in% sig_cpg_names)
    pct_sig <- if (length(sig_cpg_names)>0) 100*in_dmp/length(sig_cpg_names) else 0
    pct_cov <- if (length(sc)>0) 100*in_dmp/length(sc) else 0
    val_txt <- ""
    if (has_val) {
      vr <- find_val_result(sn, val_res)
      if (!is.null(vr)) {
        em <- extract_metrics(vr); parts <- c()
        if (!is.null(em$silhouette)) parts <- c(parts, sprintf("Sil=%.3f", em$silhouette))
        if (!is.null(em$accuracy))   parts <- c(parts, sprintf("Acc=%.1f%%", 100*em$accuracy))
        if (length(parts)>0) val_txt <- paste(parts, collapse=", ")
      }
    }
    data.frame(Signature=sn, Source=sig_sources[sn], Sig_CpGs=length(sc),
               Overlap=in_dmp, Pct_DMPs=sprintf("%.1f%%",pct_sig),
               Pct_Sig=sprintf("%.1f%%",pct_cov), Validation=val_txt,
               stringsAsFactors=FALSE)
  }))
  ovl_df <- ovl_df[order(-ovl_df$Overlap), ]
  print(knitr::kable(ovl_df, format="html", row.names=FALSE,
    table.attr='class="table table-striped table-bordered" style="font-size:12px;"'))
  cat("\n\n")
}

# ══════════════════════════════════════════════════════
# ## All DMPs {.tabset}
# ══════════════════════════════════════════════════════
cat("\n\n## All DMPs {.tabset}\n\n")

# Use df_full for all DMPs panels
render_dmp_panels(df_full, label = "All DMPs", heading = "###")

# Heatmap of top DMPs
if (length(sig_cpg_names) >= 2) {
  cat("### Heatmap\n\n")
  hm_cpgs <- head(sig_cpg_names[order(df_full[sig_cpg_names, "adj.P.Val"])], 100)
  hm_cpgs <- intersect(hm_cpgs, rownames(beta))
  if (length(hm_cpgs) >= 2) {
    bhm <- impute_row_mean(beta[hm_cpgs, , drop=FALSE])
    annot_c <- data.frame(Group=groups, row.names=colnames(beta))
    tryCatch({
      hm <- pheatmap::pheatmap(bhm, scale="row",
        show_rownames=length(hm_cpgs)<=60, show_colnames=ncol(beta)<=40,
        annotation_col=annot_c, annotation_colors=list(Group=pal),
        clustering_method="ward.D2",
        color=colorRampPalette(c("#3498db","#ecf0f1","#e74c3c"))(100),
        border_color=NA,
        fontsize_row=if(length(hm_cpgs)<=30) 7 else 5,
        fontsize_col=if(ncol(beta)<=20) 9 else 7,
        main=sprintf("Top %d DMPs", length(hm_cpgs)), silent=TRUE)
      render_plot(hm$gtable, fig.width=10, fig.height=8)
    }, error=function(e) cat("Heatmap failed.\n"))
  }
  cat("\n\n")
}

# ══════════════════════════════════════════════════════
# ## One tab per signature
# ══════════════════════════════════════════════════════
if (has_sigs) {
  for (sn in names(signatures)) {
    sig_cpgs <- signatures[[sn]]
    shared   <- intersect(sig_cpgs, rownames(df_full))

    if (length(shared) < 2) next

    df_sig_sub <- df_full[shared, , drop=FALSE]
    n_sig_in <- sum(df_sig_sub$Status != "NS")

    cat(sprintf("\n\n## %s {.tabset}\n\n", sn))

    # ── Summary sub-tab ──
    cat("### Summary\n\n")

    src <- sig_sources[sn]
    badge_col <- if (identical(src,"Default")) "#3498db" else "#27ae60"
    cat(sprintf(
      '<div class="info-box"><span style="background:%s;color:#fff;padding:2px 8px;border-radius:4px;font-size:11px;">%s</span> &nbsp; <strong>%d CpGs</strong> in signature, <strong>%d</strong> in DMP table, <strong>%d</strong> significant.</div>\n\n',
      badge_col, src, length(sig_cpgs), length(shared), n_sig_in))

    # Stat cards for this signature
    cat('<div style="display:flex;flex-wrap:wrap;gap:8px;margin:10px 0;">\n')
    cat(sig_card(length(sig_cpgs), "CpGs in Sig.", "#34495e"))
    cat(sig_card(length(shared), "In DMP Table", "#8e44ad"))
    cat(sig_card(n_sig_in, "Significant", "#e74c3c"))
    cat(sig_card(sum(df_sig_sub$Status=="Hyper"), "Hyper", "#d9534f"))
    cat(sig_card(sum(df_sig_sub$Status=="Hypo"), "Hypo", "#0275d8"))

    # Validation metrics
    if (has_val) {
      vr <- find_val_result(sn, val_res)
      if (!is.null(vr)) {
        em <- extract_metrics(vr)
        if (!is.null(em$silhouette)) {
          sc <- if (em$silhouette>0.5) "#27ae60" else if (em$silhouette>0.25) "#f39c12" else "#e74c3c"
          cat(sig_card(sprintf("%.3f", em$silhouette), "Silhouette", sc))
        }
        if (!is.null(em$accuracy)) {
          cat(sig_card(sprintf("%.1f%%", 100*em$accuracy), "Accuracy", "#8e44ad"))
        }
        if (isTRUE(em$has_svm)) {
          cat(sig_card("&#10004;", "SVM", "#9b59b6"))
        }
        if (!is.null(em$metric_used)) {
          cat(sig_card(em$metric_used, "Metric", "#1abc9c"))
        }
      }
    }
    cat('</div>\n\n')

    # Coverage of DMPs
    if (length(sig_cpg_names) > 0) {
      pct <- 100 * sum(sig_cpgs %in% sig_cpg_names) / length(sig_cpg_names)
      cat(sprintf('<div class="info-box">This signature covers <strong>%.1f%%</strong> of all significant DMPs.</div>\n\n', pct))
    }

    # ── DMP panels ──
    render_dmp_panels(df_sig_sub, label = sn, heading = "###")

    # ── Heatmap ──
    sig_in_beta <- intersect(shared, rownames(beta))
    if (length(sig_in_beta) >= 2) {
      cat("### Heatmap\n\n")
      hm_cpgs <- if (length(sig_in_beta)>150) sig_in_beta[1:150] else sig_in_beta
      bhm <- impute_row_mean(beta[hm_cpgs, , drop=FALSE])
      annot_c <- data.frame(Group=groups, row.names=colnames(beta))
      tryCatch({
        hm <- pheatmap::pheatmap(bhm, scale="row",
          show_rownames=length(hm_cpgs)<=60, show_colnames=ncol(beta)<=40,
          annotation_col=annot_c, annotation_colors=list(Group=pal),
          clustering_method="ward.D2",
          color=colorRampPalette(c("#3498db","#ecf0f1","#e74c3c"))(100),
          border_color=NA,
          fontsize_row=if(length(hm_cpgs)<=30) 7 else 5,
          fontsize_col=if(ncol(beta)<=20) 9 else 7,
          main=sprintf("%s (%d CpGs)", sn, length(hm_cpgs)), silent=TRUE)
        render_plot(hm$gtable, fig.width=10, fig.height=8)
      }, error=function(e) cat("Heatmap failed.\n"))
      cat("\n\n")
    }

    # ── PCA ──
    if (length(sig_in_beta) >= 3) {
      cat("### PCA\n\n")
      bsig <- impute_row_mean(beta[sig_in_beta, , drop=FALSE])
      pca_s <- prcomp(t(bsig), scale.=TRUE, center=TRUE)
      vp_s <- round(100*pca_s$sdev^2/sum(pca_s$sdev^2), 1)
      pdf_s <- data.frame(PC1=pca_s$x[,1], PC2=pca_s$x[,2],
                           Group=groups, Sample=colnames(beta))
      p <- ggplot(pdf_s, aes(PC1, PC2, colour=Group, label=Sample)) +
        geom_point(size=3.5, alpha=.85) + scale_colour_manual(values=pal) +
        labs(title=sprintf("PCA - %s (%d CpGs)", sn, length(sig_in_beta)),
             x=sprintf("PC1 (%s%%)", vp_s[1]), y=sprintf("PC2 (%s%%)", vp_s[2])) +
        theme_minimal(base_size=13) + theme(legend.position="bottom")
      grp_c <- table(groups)
      if (any(grp_c>=3)) {
        p <- p + stat_ellipse(level=.95, linetype=2, linewidth=.7, show.legend=FALSE,
                               data=pdf_s[groups %in% names(grp_c[grp_c>=3]),])
      }
      render_plot(p, fig.width=10, fig.height=7)
      cat("\n\n")
    }
  }
}
```

<!-- ================================================================== -->
<!-- 4. SIGNATURES & VALIDATION                                          -->
<!-- ================================================================== -->

```{r sig_gate, results='asis'}
if (!has_sigs && !has_val) {
  cat("\n# Signatures & Validation\n\n",
      '<div class="info-box">No signatures found.</div>\n')
}
```

```{r sig_content, eval=has_sigs, results='asis', fig.width=10, fig.height=7}
cat("\n# Signatures & Validation {.tabset}\n\n")

# ── Inventory ──
cat("## Inventory\n\n")
n_default <- sum(sig_sources=="Default"); n_user <- sum(sig_sources=="User")
cat('<div style="display:flex;flex-wrap:wrap;gap:8px;margin:10px 0 20px 0;">')
cat(sig_card(length(signatures), "Total", "#8e44ad"))
cat(sig_card(n_default, "Default", "#3498db"))
cat(sig_card(n_user, "User", "#27ae60"))
cat('</div>\n\n')

inv_df <- data.frame(
  Signature=names(signatures),
  Source=unname(sig_sources[names(signatures)]),
  N_CpGs=sapply(signatures, length),
  In_Beta=sapply(signatures, function(cpgs) sum(cpgs %in% rownames(beta))),
  stringsAsFactors=FALSE)
inv_df$Coverage <- sprintf("%.0f%%", 100*inv_df$In_Beta/pmax(inv_df$N_CpGs,1))

if (has_val) {
  inv_df$Silhouette <- sapply(names(signatures), function(nm) {
    v <- find_val_result(nm, val_res); if (is.null(v)) return(NA_character_)
    em <- extract_metrics(v)
    if (!is.null(em$silhouette)) sprintf("%.3f", em$silhouette) else NA_character_
  })
  inv_df$Accuracy <- sapply(names(signatures), function(nm) {
    v <- find_val_result(nm, val_res); if (is.null(v)) return(NA_character_)
    em <- extract_metrics(v)
    if (!is.null(em$accuracy)) sprintf("%.1f%%", 100*em$accuracy) else NA_character_
  })
}
print(knitr::kable(inv_df, format="html", row.names=FALSE,
  table.attr='class="table table-striped table-bordered" style="font-size:12px;"'))
cat("\n\n")

# ── Per-signature ──
for (sig_name in names(signatures)) {
  cpgs <- signatures[[sig_name]]
  cpgs_in_beta <- intersect(cpgs, rownames(beta))

  cat(sprintf("\n\n## %s {.tabset}\n\n", sig_name))

  src <- sig_sources[sig_name]
  badge_col <- if (identical(src,"Default")) "#3498db" else "#27ae60"
  cat(sprintf(
    '<div class="info-box"><span style="background:%s;color:#fff;padding:2px 8px;border-radius:4px;font-size:11px;">%s</span> &nbsp; <strong>%d CpGs</strong> (%d in beta).</div>\n\n',
    badge_col, src, length(cpgs), length(cpgs_in_beta)))

  # Metrics
  cat("### Metrics\n\n")
  v <- find_val_result(sig_name, val_res)
  if (is.null(v)) {
    cat('<div class="warn-box">No validation result found for this signature.</div>\n\n')
  } else {
    em <- extract_metrics(v)
    has_any_card <- FALSE
    cat('<div style="display:flex;flex-wrap:wrap;gap:8px;margin:10px 0;">\n')
    if (!is.null(em$n_cpgs)) { cat(sig_card(em$n_cpgs, "CpGs", "#34495e")); has_any_card <- TRUE }
    if (!is.null(em$silhouette)) {
      sc <- if (em$silhouette>0.5) "#27ae60" else if (em$silhouette>0.25) "#f39c12" else "#e74c3c"
      cat(sig_card(sprintf("%.3f", em$silhouette), "Silhouette", sc)); has_any_card <- TRUE
    }
    if (!is.null(em$accuracy)) {
      lab <- if (!is.null(em$sd_accuracy)) sprintf("%.1f%% +/- %.1f%%", 100*em$accuracy, 100*em$sd_accuracy)
             else sprintf("%.1f%%", 100*em$accuracy)
      cat(sig_card(lab, "Accuracy", "#8e44ad")); has_any_card <- TRUE
    }
    if (!is.null(em$svm_accuracy)) { cat(sig_card(sprintf("%.1f%%", 100*em$svm_accuracy), "SVM Acc.", "#8e44ad")); has_any_card <- TRUE }
    if (!is.null(em$auc)) { cat(sig_card(sprintf("%.3f", em$auc), "AUC", "#e74c3c")); has_any_card <- TRUE }
    if (!is.null(em$sensitivity)) { cat(sig_card(sprintf("%.1f%%", 100*em$sensitivity), "Sensitivity", "#3498db")); has_any_card <- TRUE }
    if (!is.null(em$specificity)) { cat(sig_card(sprintf("%.1f%%", 100*em$specificity), "Specificity", "#27ae60")); has_any_card <- TRUE }
    if (isTRUE(em$has_svm)) { cat(sig_card("&#10004;", "SVM Model", "#9b59b6")); has_any_card <- TRUE }
    if (!is.null(em$metric_used)) { cat(sig_card(em$metric_used, "Metric Used", "#1abc9c")); has_any_card <- TRUE }
    if (!is.null(em$pca_var_pct)) {
      pc12 <- sum(em$pca_var_pct[1:min(2,length(em$pca_var_pct))])
      cat(sig_card(sprintf("%.1f%%", pc12), "PCA Var (PC1+2)", "#e67e22")); has_any_card <- TRUE
    }
    cat('</div>\n\n')
    if (!has_any_card) cat('<div class="warn-box">No displayable metrics.</div>\n\n')

    # Silhouette interpretation
    if (!is.null(em$silhouette)) {
      sil_v <- em$silhouette
      interp <- if (sil_v>0.7) "Strong structure" else if (sil_v>0.5) "Reasonable structure"
                else if (sil_v>0.25) "Weak structure" else "No substantial structure"
      cat(sprintf('<div class="info-box"><strong>Silhouette = %.3f</strong> &mdash; %s. Values &gt; 0.5 indicate good cluster separation.</div>\n\n', sil_v, interp))
      if (!is.null(em$silhouette_per_cluster)) {
        scd <- data.frame(Cluster=factor(seq_along(em$silhouette_per_cluster)),
                          Sil=em$silhouette_per_cluster)
        p <- ggplot(scd, aes(Cluster, Sil, fill=Sil>0.5)) +
          geom_col(alpha=.85, show.legend=FALSE) +
          scale_fill_manual(values=c("TRUE"="#27ae60","FALSE"="#e74c3c")) +
          geom_hline(yintercept=em$silhouette, linetype="dashed", colour="#3498db") +
          labs(title="Per-Cluster Silhouette", x="Cluster", y="Avg. Silhouette") +
          theme_minimal(base_size=12)
        render_plot(p, fig.width=8, fig.height=4); cat("\n\n")
      }
    }

    # SVM
    if (isTRUE(em$has_svm)) {
      cat('<div class="info-box"><strong>SVM Classifier</strong>')
      if (!is.null(em$svm_kernel)) cat(sprintf(' &mdash; Kernel: %s', em$svm_kernel))
      if (!is.null(em$svm_nSV)) cat(sprintf(' &mdash; SVs: %s', paste(em$svm_nSV, collapse=" / ")))
      cat('</div>\n\n')
      if (!is.null(em$svm_cv_accuracies) && length(em$svm_cv_accuracies)>1) {
        fdf <- data.frame(Fold=factor(seq_along(em$svm_cv_accuracies)), Acc=em$svm_cv_accuracies)
        p <- ggplot(fdf, aes(Fold, Acc)) + geom_col(fill="#9b59b6", alpha=.85) +
          geom_hline(yintercept=mean(em$svm_cv_accuracies), linetype="dashed", colour="#e74c3c") +
          scale_y_continuous(limits=c(0,1), labels=function(x) paste0(x*100,"%")) +
          labs(title="SVM CV Accuracy", x="Fold", y="Accuracy") + theme_minimal(base_size=12)
        render_plot(p, fig.width=10, fig.height=5); cat("\n\n")
      }
    }

    # Fold accuracies
    if (!is.null(em$fold_accuracies) && length(em$fold_accuracies)>1) {
      fdf <- data.frame(Fold=factor(seq_along(em$fold_accuracies)), Acc=em$fold_accuracies)
      p <- ggplot(fdf, aes(Fold, Acc)) + geom_col(fill="#3498db", alpha=.85) +
        geom_hline(yintercept=mean(em$fold_accuracies), linetype="dashed", colour="#e74c3c") +
        scale_y_continuous(limits=c(0,1), labels=function(x) paste0(x*100,"%")) +
        labs(title=paste0(sig_name," - Per-Fold Accuracy"), x="Fold", y="Accuracy") +
        theme_minimal(base_size=12)
      render_plot(p, fig.width=10, fig.height=5); cat("\n\n")
    }

    # Confusion
    if (!is.null(em$confusion)) {
      cat("**Confusion Matrix:**\n\n")
      print(knitr::kable(as.data.frame.matrix(em$confusion), format="html",
        table.attr='class="table table-bordered" style="width:auto;"'))
      cat("\n\n")
    }

    # ROC
    if (!is.null(em$roc) && has_proc) {
      tryCatch({
        rdf <- data.frame(FPR=1-em$roc$specificities, TPR=em$roc$sensitivities)
        auc_val <- as.numeric(pROC::auc(em$roc))
        p <- ggplot(rdf, aes(FPR, TPR)) + geom_line(colour="#e74c3c", linewidth=1.2) +
          geom_abline(slope=1, intercept=0, linetype="dashed", colour="grey60") +
          annotate("text", x=.6, y=.3, size=5, fontface="bold", label=sprintf("AUC=%.3f",auc_val)) +
          labs(title=sprintf("ROC - %s", sig_name), x="1-Specificity", y="Sensitivity") +
          theme_minimal(base_size=13)
        render_plot(p, fig.width=8, fig.height=6)
      }, error=function(e) cat("ROC plot failed.\n"))
      cat("\n\n")
    }

    # PCA variance from validation
    if (!is.null(em$pca_var_pct)) {
      n_show <- min(10, length(em$pca_var_pct))
      sdf <- data.frame(PC=seq_len(n_show), V=em$pca_var_pct[seq_len(n_show)])
      p <- ggplot(sdf, aes(PC, V)) + geom_col(fill="#e67e22", alpha=.8) +
        geom_line(colour="#e74c3c", linewidth=1) + geom_point(colour="#e74c3c", size=2) +
        labs(title=sprintf("PCA Variance - %s", sig_name), x="PC", y="Variance (%%)") +
        scale_x_continuous(breaks=sdf$PC) + theme_minimal(base_size=12)
      render_plot(p, fig.width=8, fig.height=4); cat("\n\n")
    }
  }

  # PCA from beta
  if (length(cpgs_in_beta)>=3) {
    cat("### PCA\n\n")
    bsig <- impute_row_mean(beta[cpgs_in_beta,,drop=FALSE])
    pca_s <- prcomp(t(bsig), scale.=TRUE, center=TRUE)
    vp_s <- round(100*pca_s$sdev^2/sum(pca_s$sdev^2), 1)
    pdf_s <- data.frame(PC1=pca_s$x[,1], PC2=pca_s$x[,2], Group=groups, Sample=colnames(beta))
    p <- ggplot(pdf_s, aes(PC1, PC2, colour=Group, label=Sample)) +
      geom_point(size=3.5, alpha=.85) + scale_colour_manual(values=pal) +
      labs(title=sprintf("PCA - %s (%d CpGs)", sig_name, length(cpgs_in_beta)),
           x=sprintf("PC1 (%s%%)", vp_s[1]), y=sprintf("PC2 (%s%%)", vp_s[2])) +
      theme_minimal(base_size=13) + theme(legend.position="bottom")
    grp_c <- table(groups)
    if (any(grp_c>=3)) {
      p <- p + stat_ellipse(level=.95, linetype=2, linewidth=.7, show.legend=FALSE,
                             data=pdf_s[groups %in% names(grp_c[grp_c>=3]),])
    }
    render_plot(p, fig.width=10, fig.height=7); cat("\n\n")
  } else { cat("### PCA\n\nFewer than 3 CpGs in beta.\n\n") }

  # Heatmap
  if (length(cpgs_in_beta)>=2) {
    cat("### Heatmap\n\n")
    hm_cpgs <- if (length(cpgs_in_beta)>150) cpgs_in_beta[1:150] else cpgs_in_beta
    bhm <- impute_row_mean(beta[hm_cpgs,,drop=FALSE])
    annot_c <- data.frame(Group=groups, row.names=colnames(beta))
    tryCatch({
      hm <- pheatmap::pheatmap(bhm, scale="row",
        show_rownames=length(hm_cpgs)<=60, show_colnames=ncol(beta)<=40,
        annotation_col=annot_c, annotation_colors=list(Group=pal),
        clustering_method="ward.D2",
        color=colorRampPalette(c("#3498db","#ecf0f1","#e74c3c"))(100),
        border_color=NA,
        fontsize_row=if(length(hm_cpgs)<=30) 7 else 5,
        fontsize_col=if(ncol(beta)<=20) 9 else 7,
        main=sprintf("%s - %d CpGs", sig_name, length(hm_cpgs)), silent=TRUE)
      render_plot(hm$gtable, fig.width=10, fig.height=8)
    }, error=function(e) cat("Heatmap failed.\n"))
    cat("\n\n")
  } else { cat("### Heatmap\n\nFewer than 2 CpGs.\n\n") }
}
```

<!-- ================================================================== -->
<!-- FOOTER                                                              -->
<!-- ================================================================== -->

---

<div style="text-align:center; color:#999; margin:30px 0;">
Generated by <strong>mETHYLotest</strong> on `r format(Sys.time(), "%d %B %Y at %H:%M")`
</div>
