---
title: "mETHYLotest NGS Signature Validation Report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
params:
  data_path: "data.rds"
---

```{=html}
<style>
.metric-box {
  background-color: #f5f5f5;
  border: 1px solid #e3e3e3;
  border-radius: 4px;
  padding: 15px;
  text-align: center;
  margin-bottom: 20px;
}
.metric-title { font-weight: bold; color: #555; font-size: 1.1em; }
.metric-value { font-size: 2em; font-weight: bold; margin-top: 5px; }
.col-blue   { color: #2980b9; }
.col-green  { color: #27ae60; }
.col-orange { color: #f39c12; }
.col-red    { color: #e74c3c; }
.col-purple { color: #8e44ad; }
.section-sep { margin-top: 50px; border-top: 2px solid #ddd; padding-top: 20px; }
.sig-header { background: #f8f9fa; padding: 12px 15px; border-radius: 5px;
              border-left: 4px solid #2980b9; margin-bottom: 15px; }
.warn-box { background: #fff3cd; border: 1px solid #ffc107; border-radius: 4px;
            padding: 12px; margin: 10px 0; }
.good-box { background: #d5f5e3; border: 1px solid #27ae60; border-radius: 4px;
            padding: 12px; margin: 10px 0; }
.bad-box  { background: #fdedec; border: 1px solid #e74c3c; border-radius: 4px;
            padding: 12px; margin: 10px 0; }
.info-box { background: #eaf2f8; border: 1px solid #3498db; border-radius: 4px;
            padding: 12px; margin: 10px 0; }
.cpg-table { font-size: 0.85em; }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 7,
  dpi = 100
)

library(ggplot2)
library(plotly)
library(caret)
library(pROC)
library(htmltools)
library(knitr)

has_heatmaply <- requireNamespace("heatmaply", quietly = TRUE)
has_pheatmap  <- requireNamespace("pheatmap",  quietly = TRUE)
has_cluster   <- requireNamespace("cluster",   quietly = TRUE)

results <- readRDS(params$data_path)
```

# Validation Overview

This report benchmarks **`r length(results)`** epigenetic signature(s) from **NGS** data:

- **Unsupervised**: PCA and hierarchical clustering to assess natural group separation.
- **Supervised**: SVM with cross-validation (LOOCV or k-fold) to estimate predictive performance.
- **Clustering quality**: Silhouette score to quantify separation strength.

Signatures prefixed with `DMC_default_` were generated automatically from significant DMCs
(top sites by combined score). Other signatures were created manually via the DiffMeth Explorer UI.

Loci are identified by genomic coordinates (`chr:start-end`), not probe IDs.

# Benchmark Summary

```{r benchmark_table}
bench_rows <- lapply(names(results), function(nm) {
  r <- results[[nm]]
  svm_ok <- !is.null(r$svm_model)

  perf <- if (svm_ok) {
    round(max(r$svm_model$results[[r$metric_used]], na.rm = TRUE), 3)
  } else { NA }

  metric2_val <- NA
  metric2_lbl <- "Secondary"
  if (svm_ok) {
    if (r$metric_used == "ROC" && "Sens" %in% colnames(r$svm_model$results)) {
      metric2_val <- round(max(r$svm_model$results$Sens, na.rm = TRUE), 3)
      metric2_lbl <- "Sensitivity"
    } else if ("Kappa" %in% colnames(r$svm_model$results)) {
      metric2_val <- round(max(r$svm_model$results$Kappa, na.rm = TRUE), 3)
      metric2_lbl <- "Kappa"
    }
  }

  # Direct metrics from validate()
  acc_str <- if (!is.null(r$accuracy) && !is.na(r$accuracy)) {
    sprintf("%.1f%%", r$accuracy * 100)
  } else { "N/A" }

  is_default <- grepl("^DMC_default_", nm)

  data.frame(
    Signature   = nm,
    Source      = if (is_default) "Auto (DMC)" else "User",
    CpGs        = r$n_cpgs,
    Metric      = if (svm_ok) r$metric_used else "N/A",
    Performance = perf,
    Accuracy    = acc_str,
    Secondary   = metric2_val,
    Silhouette  = round(r$silhouette, 3),
    stringsAsFactors = FALSE
  )
})

bench_df <- do.call(rbind, bench_rows)
bench_df <- bench_df[order(-bench_df$Performance, na.last = TRUE), ]

knitr::kable(bench_df, row.names = FALSE,
             caption = "All signatures ranked by primary metric")
```

```{r benchmark_plot}
bench_plot <- bench_df[!is.na(bench_df$Performance), ]

if (nrow(bench_plot) > 0) {
  bench_plot$Signature <- factor(bench_plot$Signature,
                                  levels = bench_plot$Signature[
                                    order(bench_plot$Performance)])

  p <- ggplot(bench_plot,
              aes(x = Signature, y = Performance, fill = Source)) +
    geom_bar(stat = "identity", alpha = 0.85) +
    geom_text(aes(label = Performance), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_fill_manual(values = c("Auto (DMC)" = "#2980b9",
                                  "User" = "#27ae60")) +
    labs(title = "Signature Performance Comparison",
         x = "", y = "Performance (AUC / Accuracy)") +
    theme_minimal() +
    ylim(0, max(bench_plot$Performance, na.rm = TRUE) * 1.15)

  ggplotly(p)
}
```

```{r silhouette_comparison}
sil_df <- bench_df[, c("Signature", "Source", "Silhouette")]
sil_df$Signature <- factor(sil_df$Signature,
                             levels = sil_df$Signature[order(sil_df$Silhouette)])
sil_df$Quality <- ifelse(sil_df$Silhouette > 0.5, "Good",
                  ifelse(sil_df$Silhouette > 0.25, "Weak", "Poor"))

p_sil <- ggplot(sil_df, aes(x = Signature, y = Silhouette, fill = Quality)) +
  geom_bar(stat = "identity", alpha = 0.85) +
  geom_text(aes(label = Silhouette), hjust = -0.1, size = 3.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "#27ae60") +
  geom_hline(yintercept = 0.25, linetype = "dotted", colour = "#f39c12") +
  coord_flip() +
  scale_fill_manual(values = c("Good" = "#27ae60", "Weak" = "#f39c12", "Poor" = "#e74c3c")) +
  labs(title = "Silhouette Score Comparison",
       x = "", y = "Silhouette") +
  theme_minimal() +
  ylim(0, max(sil_df$Silhouette, na.rm = TRUE) * 1.2)

ggplotly(p_sil)
```

# Individual Signature Reports

```{r individual_reports}
doc_content <- list()

for (sig_name in names(results)) {

  res    <- results[[sig_name]]
  svm_ok <- !is.null(res$svm_model)
  is_default   <- grepl("^DMC_default_", sig_name)
  source_label <- if (is_default) "Automatic (from DMC)" else "User-created"

  # ── Header ──
  doc_content[[length(doc_content) + 1]] <- div(class = "section-sep")
  doc_content[[length(doc_content) + 1]] <- div(
    class = "sig-header",
    h2(sig_name, style = "margin:0;"),
    p(style = "margin:5px 0 0 0; color:#666;",
      paste0(res$n_cpgs, " CpGs | Source: ", source_label,
             " | Metric: ", if (svm_ok) res$metric_used else "N/A"))
  )

  # ── Metrics boxes ──
  val_m1  <- "FAILED"
  lbl_m1  <- "SVM Result"
  val_m2  <- "N/A"
  lbl_m2  <- "Secondary"
  css_m1  <- "col-red"

  if (svm_ok) {
    if (res$metric_used == "ROC") {
      lbl_m1 <- "AUC (ROC)"
      val_m1 <- round(max(res$svm_model$results$ROC, na.rm = TRUE), 3)
      if ("Sens" %in% colnames(res$svm_model$results)) {
        lbl_m2 <- "Sensitivity"
        val_m2 <- round(max(res$svm_model$results$Sens, na.rm = TRUE), 3)
      }
    } else {
      lbl_m1 <- "Accuracy"
      val_m1 <- round(max(res$svm_model$results$Accuracy, na.rm = TRUE), 3)
      if ("Kappa" %in% colnames(res$svm_model$results)) {
        lbl_m2 <- "Kappa"
        val_m2 <- round(max(res$svm_model$results$Kappa, na.rm = TRUE), 3)
      }
    }
    css_m1 <- "col-blue"
  }

  # Direct accuracy if available
  val_acc <- "N/A"
  if (!is.null(res$accuracy) && !is.na(res$accuracy)) {
    val_acc <- sprintf("%.1f%%", res$accuracy * 100)
  }

  val_sil <- round(res$silhouette, 3)
  css_sil <- if (val_sil > 0.5) "col-green"
             else if (val_sil > 0.25) "col-orange"
             else "col-red"

  doc_content[[length(doc_content) + 1]] <- div(
    class = "row",
    div(class = "col-sm-3",
        div(class = "metric-box",
            div(class = "metric-title", lbl_m1),
            div(class = paste("metric-value", css_m1), val_m1))),
    div(class = "col-sm-3",
        div(class = "metric-box",
            div(class = "metric-title", lbl_m2),
            div(class = "metric-value col-green", val_m2))),
    div(class = "col-sm-3",
        div(class = "metric-box",
            div(class = "metric-title", "Accuracy"),
            div(class = "metric-value col-purple", val_acc))),
    div(class = "col-sm-3",
        div(class = "metric-box",
            div(class = "metric-title", "Silhouette"),
            div(class = paste("metric-value", css_sil), val_sil)))
  )

  # Silhouette interpretation
  sil_interp <- if (val_sil > 0.7) list("Strong cluster separation.", "good-box")
                else if (val_sil > 0.5) list("Reasonable structure.", "info-box")
                else if (val_sil > 0.25) list("Weak structure.", "warn-box")
                else list("No substantial structure.", "bad-box")

  doc_content[[length(doc_content) + 1]] <- div(
    class = sil_interp[[2]],
    strong(paste0("Silhouette = ", val_sil)),
    " -- ", sil_interp[[1]]
  )

  # Confusion matrix (if available directly)
  if (!is.null(res$confusion)) {
    doc_content[[length(doc_content) + 1]] <- div(
      class = "info-box",
      strong("Confusion Matrix (from CV predictions):"),
      br(),
      tags$table(
        class = "table table-bordered",
        style = "width:auto; margin-top:8px;",
        tags$thead(tags$tr(
          tags$th(""),
          lapply(colnames(res$confusion), tags$th)
        )),
        tagList(lapply(seq_len(nrow(res$confusion)), function(i) {
          tags$tr(
            tags$td(strong(rownames(res$confusion)[i])),
            tagList(lapply(seq_len(ncol(res$confusion)), function(j) {
              tags$td(res$confusion[i, j])
            }))
          )
        }))
      )
    )
  }

  # ── 1. PCA ──
  doc_content[[length(doc_content) + 1]] <- h3("1. PCA (Unsupervised Separation)")

  if (!is.null(res$pca)) {
    pca    <- res$pca
    df_pca <- data.frame(
      PC1    = pca$x[, 1],
      PC2    = pca$x[, 2],
      Group  = res$raw_data$Class,
      Sample = rownames(res$raw_data))
    ve <- round(summary(pca)$importance[2, 1:2] * 100, 1)

    p_pca <- ggplot(df_pca, aes(x = PC1, y = PC2,
                                 color = Group, text = Sample)) +
      geom_point(size = 3, alpha = 0.8) +
      stat_ellipse(level = 0.95, linetype = "dashed", alpha = 0.4) +
      labs(x = paste0("PC1 (", ve[1], "%)"),
           y = paste0("PC2 (", ve[2], "%)"),
           title = paste0("PCA - ", sig_name)) +
      theme_minimal() +
      scale_color_manual(values = c("CTL" = "#3498db", "Test" = "#e74c3c"))

    doc_content[[length(doc_content) + 1]] <- ggplotly(p_pca, tooltip = "text")

    # Scree plot
    ve_all  <- summary(pca)$importance[2, ] * 100
    n_show  <- min(10, length(ve_all))
    df_scree <- data.frame(PC = seq_len(n_show), Var = ve_all[seq_len(n_show)])

    p_scree <- ggplot(df_scree, aes(PC, Var)) +
      geom_col(fill = "#2980b9", alpha = 0.8) +
      geom_line(colour = "#e74c3c", linewidth = 1) +
      geom_point(colour = "#e74c3c", size = 2) +
      scale_x_continuous(breaks = df_scree$PC) +
      labs(title = "Scree Plot", x = "PC", y = "Variance (%)") +
      theme_minimal()

    doc_content[[length(doc_content) + 1]] <- ggplotly(p_scree)

  } else {
    doc_content[[length(doc_content) + 1]] <- p("PCA unavailable for this signature.")
  }

  # ── 2. Heatmap ──
  doc_content[[length(doc_content) + 1]] <- h3("2. Hierarchical Clustering")

  if (!is.null(res$beta_matrix)) {
    bm <- res$beta_matrix
    # Limit for heatmaply performance
    if (nrow(bm) > 200) {
      rv <- apply(bm, 1, var, na.rm = TRUE)
      bm <- bm[order(rv, decreasing = TRUE)[1:200], , drop = FALSE]
    }

    annot_col <- data.frame(Group = res$raw_data$Class)
    rownames(annot_col) <- rownames(res$raw_data)

    if (has_heatmaply) {
      p_heat <- tryCatch({
        heatmaply::heatmaply(
          bm,
          limits          = c(0, 1),
          colors          = colorRampPalette(c("#2980b9", "white", "#c0392b"))(100),
          col_side_colors = annot_col,
          row_dend_left   = TRUE,
          col_dend_top    = TRUE,
          branches_lwd    = 0.2,
          labRow = if (nrow(bm) > 50) NULL else rownames(bm),
          main   = paste0("Methylation - ", nrow(bm), " CpGs"),
          xlab   = "Samples",
          ylab   = "CpG Loci"
        )
      }, error = function(e) NULL)

      if (!is.null(p_heat)) {
        doc_content[[length(doc_content) + 1]] <- p_heat
      } else {
        doc_content[[length(doc_content) + 1]] <- p("Heatmap rendering failed.")
      }
    } else {
      doc_content[[length(doc_content) + 1]] <- p("heatmaply package not installed.")
    }
  } else {
    doc_content[[length(doc_content) + 1]] <- p("Beta matrix not available.")
  }

  # ── 3. SVM Performance ──
  doc_content[[length(doc_content) + 1]] <- h3("3. SVM Performance")

  if (!svm_ok) {
    doc_content[[length(doc_content) + 1]] <- div(
      class = "warn-box",
      strong("SVM analysis not possible."),
      br(),
      "Sample size too small or classes too unbalanced.",
      " Rely on PCA and Heatmap for qualitative assessment."
    )
  } else {
    # Confusion matrix plot
    preds <- res$svm_model$pred
    cm_plot <- NULL
    if (!is.null(preds)) {
      if ("C" %in% colnames(preds) && !is.null(res$svm_model$bestTune$C)) {
        preds <- preds[preds$C == res$svm_model$bestTune$C, ]
      }
      if (nrow(preds) > 0) {
        cm    <- confusionMatrix(preds$pred, preds$obs)
        df_cm <- as.data.frame(cm$table)

        cm_plot <- ggplot(df_cm, aes(x = Reference, y = Prediction, fill = Freq)) +
          geom_tile() +
          geom_text(aes(label = Freq), size = 6, color = "white") +
          scale_fill_gradient(low = "#868e96", high = "#2980b9") +
          labs(title = "Confusion Matrix") +
          theme_minimal()
      }
    }

    # ROC
    p_roc <- NULL
    if (res$metric_used == "ROC" && !is.null(preds) && nrow(preds) > 0) {
      lvls <- levels(preds$obs)
      if (length(lvls) >= 2 && lvls[1] %in% colnames(preds)) {
        roc_obj <- tryCatch(
          pROC::roc(preds$obs, preds[[lvls[1]]]),
          error = function(e) NULL)

        if (!is.null(roc_obj)) {
          roc_df  <- data.frame(
            Specificity = roc_obj$specificities,
            Sensitivity = roc_obj$sensitivities)
          auc_val <- round(roc_obj$auc, 3)

          p_roc <- ggplot(roc_df,
                          aes(x = 1 - Specificity, y = Sensitivity)) +
            geom_line(color = "#e74c3c", linewidth = 1) +
            geom_abline(intercept = 0, slope = 1,
                        linetype = "dashed", color = "grey") +
            labs(title = paste0("ROC Curve (AUC = ", auc_val, ")")) +
            theme_minimal()
        }
      }
    }

    if (is.null(p_roc)) {
      p_roc <- ggplot() +
        annotate("text", x = 0.5, y = 0.5,
                 label = "ROC unavailable (Accuracy mode)") +
        theme_void()
    }

    if (is.null(cm_plot)) {
      cm_plot <- ggplot() +
        annotate("text", x = 0.5, y = 0.5,
                 label = "Confusion matrix unavailable") +
        theme_void()
    }

    doc_content[[length(doc_content) + 1]] <- div(
      class = "row",
      div(class = "col-sm-6", ggplotly(cm_plot)),
      div(class = "col-sm-6", ggplotly(p_roc))
    )
  }

  # ── 4. Variable Importance ──
  doc_content[[length(doc_content) + 1]] <- h3("4. Top Discriminative Loci")

  if (!svm_ok) {
    doc_content[[length(doc_content) + 1]] <- p(
      "Requires a valid SVM model. See above.")
  } else {
    imp_obj <- tryCatch(
      varImp(res$svm_model, scale = FALSE),
      error = function(e) NULL)

    if (!is.null(imp_obj) && nrow(imp_obj$importance) > 0) {
      imp <- imp_obj$importance

      imp$Score <- if ("Overall" %in% colnames(imp)) imp$Overall
                   else apply(imp, 1, max)

      imp$RawName <- rownames(imp)
      known <- names(res$genes_map)

      # Recover locus names (caret mangles chr:start-end)
      recover_name <- function(x) {
        if (x %in% known) return(x)
        # caret: "chr1.12345.12345" or "X.chr1.12345.12345"
        x2 <- gsub("\\.", ":", x)
        x2 <- sub(":([0-9]+):([0-9]+)$", ":\\1-\\2", x2)
        if (x2 %in% known) return(x2)
        x3 <- gsub("^X:", "", x2)
        if (x3 %in% known) return(x3)
        x3 <- gsub("^X", "", x)
        x3 <- gsub("\\.", ":", x3)
        x3 <- sub(":([0-9]+):([0-9]+)$", ":\\1-\\2", x3)
        if (x3 %in% known) return(x3)
        return(x)
      }
      imp$Locus <- vapply(imp$RawName, recover_name, character(1))

      # Parse chr:start for display
      parts <- strsplit(imp$Locus, "[:|-]")
      imp$Chr   <- sapply(parts, function(x) if (length(x) >= 1) x[1] else "?")
      imp$Start <- sapply(parts, function(x) if (length(x) >= 2) x[2] else "?")

      imp$Label <- imp$Locus

      plot_data <- imp[order(imp$Score, decreasing = TRUE), ]
      plot_data <- head(plot_data, 20)
      plot_data$Label <- factor(plot_data$Label,
                                levels = rev(plot_data$Label))

      p_imp <- ggplot(plot_data, aes(x = Label, y = Score)) +
        geom_bar(stat = "identity", fill = "#c0392b", alpha = 0.8) +
        coord_flip() +
        labs(title = "Top 20 Discriminative Loci",
             x = "", y = "Variable Importance") +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 9))

      doc_content[[length(doc_content) + 1]] <- ggplotly(p_imp)

      # Table
      tbl_data <- plot_data[, c("Locus", "Chr", "Start", "Score")]
      tbl_data$Score <- round(tbl_data$Score, 4)
      tbl_data$Rank  <- seq_len(nrow(tbl_data))
      tbl_data <- tbl_data[, c("Rank", "Locus", "Chr", "Start", "Score")]

      doc_content[[length(doc_content) + 1]] <- div(
        style = "margin-top:15px;",
        HTML(kable(tbl_data, row.names = FALSE,
              format = "html",
              table.attr = 'class="table table-striped cpg-table"'))
      )

    } else {
      doc_content[[length(doc_content) + 1]] <- p(
        "Variable importance could not be calculated.")
    }
  }

  # ── 5. CpG Locus Details ──
  doc_content[[length(doc_content) + 1]] <- h3("5. CpG Locus Details")

  if (!is.null(res$beta_matrix)) {
    bm <- res$beta_matrix
    groups <- res$raw_data$Class
    grp_levels <- levels(groups)

    detail <- data.frame(Locus = rownames(bm), stringsAsFactors = FALSE)

    for (g in grp_levels) {
      cols <- which(groups == g)
      if (length(cols) > 1) {
        detail[[paste0("Mean_", g)]] <- round(rowMeans(bm[, cols, drop = FALSE], na.rm = TRUE), 4)
        detail[[paste0("SD_", g)]]   <- round(apply(bm[, cols, drop = FALSE], 1, sd, na.rm = TRUE), 4)
      } else if (length(cols) == 1) {
        detail[[paste0("Mean_", g)]] <- round(bm[, cols], 4)
        detail[[paste0("SD_", g)]]   <- NA
      }
    }

    if (length(grp_levels) == 2) {
      detail$Delta <- round(
        detail[[paste0("Mean_", grp_levels[2])]] -
        detail[[paste0("Mean_", grp_levels[1])]], 4)
      detail <- detail[order(-abs(detail$Delta)), ]
    }

    # Parse coordinates
    parts <- strsplit(detail$Locus, "[:|-]")
    detail$Chr   <- sapply(parts, function(x) x[1])
    detail$Start <- as.integer(sapply(parts, function(x) if (length(x) >= 2) x[2] else NA))

    # Show top 50
    detail_show <- head(detail, 50)

    doc_content[[length(doc_content) + 1]] <- div(
      class = "info-box",
      paste0("Showing top ", nrow(detail_show), " of ", nrow(detail),
             " loci (sorted by |Delta|).")
    )

    doc_content[[length(doc_content) + 1]] <- div(
      HTML(kable(detail_show, row.names = FALSE,
            format = "html",
            table.attr = 'class="table table-striped table-condensed cpg-table"'))
    )
  } else {
    doc_content[[length(doc_content) + 1]] <- p("Beta matrix not available.")
  }

  # ── Fold Accuracies ──
  if (!is.null(res$fold_accuracies) && length(res$fold_accuracies) > 1) {
    doc_content[[length(doc_content) + 1]] <- h3("6. Per-Fold Accuracy")

    fdf <- data.frame(
      Fold = factor(seq_along(res$fold_accuracies)),
      Accuracy = res$fold_accuracies)

    p_fold <- ggplot(fdf, aes(Fold, Accuracy)) +
      geom_col(fill = "#3498db", alpha = 0.85) +
      geom_hline(yintercept = mean(res$fold_accuracies), linetype = "dashed",
                 colour = "#e74c3c", linewidth = 0.8) +
      scale_y_continuous(limits = c(0, 1),
                         labels = function(x) paste0(x*100, "%")) +
      labs(title = paste0(sig_name, " - Per-Fold Accuracy"),
           x = "Fold", y = "Accuracy") +
      theme_minimal()

    doc_content[[length(doc_content) + 1]] <- ggplotly(p_fold)
  }

  doc_content[[length(doc_content) + 1]] <- hr()
}

tagList(doc_content)
```

# Session Info

```{r session}
sessionInfo()
```
