---
title: "mETHYLotest Epigenetic Signature Validation Report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
params:
  data_path: "data.rds"
---

```{=html}
<style>
.metric-box {
  background-color: #f5f5f5;
  border: 1px solid #e3e3e3;
  border-radius: 4px;
  padding: 15px;
  text-align: center;
  margin-bottom: 20px;
}
.metric-title { font-weight: bold; color: #555; font-size: 1.1em; }
.metric-value { font-size: 2em; font-weight: bold; margin-top: 5px; }
.col-blue   { color: #0275d8; }
.col-green  { color: #5cb85c; }
.col-orange { color: #f0ad4e; }
.col-red    { color: #d9534f; }
.section-sep { margin-top: 50px; border-top: 2px solid #ddd; padding-top: 20px; }
.sig-header { background: #f8f9fa; padding: 12px 15px; border-radius: 5px;
              border-left: 4px solid #8e44ad; margin-bottom: 15px; }
.warn-box { background: #fff3cd; border: 1px solid #ffc107; border-radius: 4px;
            padding: 12px; margin: 10px 0; }
</style>
```

```{r}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 7,
  dpi = 100
)

library(ggplot2)
library(plotly)
library(heatmaply)
library(caret)
library(dplyr)
library(pROC)
library(htmltools)
library(knitr)

results <- readRDS(params$data_path)
```

# Validation Overview
This report benchmarks r`length(results)` epigenetic signature(s):

- **Unsupervised**: PCA and hierarchical clustering to assess natural group separation.
- **Supervised**: SVM with cross-validation (LOOCV or k-fold) to estimate predictive performance.
- **Clustering quality**: Silhouette score to quantify separation strength.
Signatures prefixed with `DMP_default_` were generated automatically from significant DMPs
(top 500 by combined score). Other signatures were created manually via the DMP Explorer UI.

# Benchmark Summary

```{r}
bench_rows <- lapply(names(results), function(nm) {
  r <- results[[nm]]
  svm_ok <- !is.null(r$svm_model)

  perf <- if (svm_ok)
    round(max(r$svm_model$results[[r$metric_used]], na.rm = TRUE), 3)
  else NA

  metric2_val <- NA
  if (svm_ok) {
    if (r$metric_used == "ROC") {
      metric2_val <- round(max(r$svm_model$results$Sens, na.rm = TRUE), 3)
    } else {
      metric2_val <- round(max(r$svm_model$results$Kappa, na.rm = TRUE), 3)
    }
  }

  is_default <- grepl("^DMP_default_", nm)

  data.frame(
    Signature   = nm,
    Source      = if (is_default) "Auto (DMP)" else "User",
    CpGs        = r$n_cpgs,
    Metric      = if (svm_ok) r$metric_used else "N/A",
    Performance = perf,
    Secondary   = metric2_val,
    Silhouette  = round(r$silhouette, 3),
    stringsAsFactors = FALSE
  )
})

bench_df <- do.call(rbind, bench_rows)
bench_df <- bench_df[order(-bench_df$Performance, na.last = TRUE), ]

knitr::kable(bench_df, row.names = FALSE,
             caption = "All signatures ranked by primary metric")
```

```{r}
bench_plot <- bench_df[!is.na(bench_df$Performance), ]

if (nrow(bench_plot) > 0) {
  bench_plot$Signature <- factor(bench_plot$Signature,
                                  levels = bench_plot$Signature[
                                    order(bench_plot$Performance)])

  p <- ggplot(bench_plot,
              aes(x = Signature, y = Performance, fill = Source)) +
    geom_bar(stat = "identity", alpha = 0.85) +
    geom_text(aes(label = Performance), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_fill_manual(values = c("Auto (DMP)" = "#8e44ad",
                                  "User" = "#27ae60")) +
    labs(title = "Signature Performance Comparison",
         x = "", y = "Performance (AUC / Accuracy)") +
    theme_minimal() +
    ylim(0, max(bench_plot$Performance, na.rm = TRUE) * 1.15)

  ggplotly(p)
}
```

# Individual Signature Reports

```{r}
doc_content <- list()

for (sig_name in names(results)) {

  res    <- results[[sig_name]]
  svm_ok <- !is.null(res$svm_model)
  is_default <- grepl("^DMP_default_", sig_name)
  source_label <- if (is_default) "Automatic (from DMP)" else "User-created"

  # в”Ђв”Ђ Header в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  doc_content[[length(doc_content) + 1]] <- div(class = "section-sep")
  doc_content[[length(doc_content) + 1]] <- div(
    class = "sig-header",
    h2(sig_name, style = "margin:0;"),
    p(style = "margin:5px 0 0 0; color:#666;",
      paste0(res$n_cpgs, " CpGs | Source: ", source_label,
             " | Metric: ", if (svm_ok) res$metric_used else "N/A"))
  )

  # в”Ђв”Ђ Metrics в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  val_m1    <- "FAILED"
  lbl_m1    <- "SVM Result"
  val_m2    <- "N/A"
  lbl_m2    <- "Secondary"
  css_m1    <- "col-red"

  if (svm_ok) {
    if (res$metric_used == "ROC") {
      lbl_m1 <- "ROC (AUC)"
      val_m1 <- round(max(res$svm_model$results$ROC, na.rm = TRUE), 3)
      lbl_m2 <- "Sensitivity"
      val_m2 <- round(max(res$svm_model$results$Sens, na.rm = TRUE), 3)
    } else {
      lbl_m1 <- "Accuracy"
      val_m1 <- round(max(res$svm_model$results$Accuracy, na.rm = TRUE), 3)
      lbl_m2 <- "Kappa"
      val_m2 <- round(max(res$svm_model$results$Kappa, na.rm = TRUE), 3)
    }
    css_m1 <- "col-blue"
  }

  val_sil    <- round(res$silhouette, 3)
  css_sil    <- if (val_sil > 0.5) "col-green"
                else if (val_sil > 0.25) "col-orange"
                else "col-red"

  doc_content[[length(doc_content) + 1]] <- div(
    class = "row",
    div(class = "col-sm-4",
        div(class = "metric-box",
            div(class = "metric-title", lbl_m1),
            div(class = paste("metric-value", css_m1), val_m1))),
    div(class = "col-sm-4",
        div(class = "metric-box",
            div(class = "metric-title", lbl_m2),
            div(class = "metric-value col-green", val_m2))),
    div(class = "col-sm-4",
        div(class = "metric-box",
            div(class = "metric-title", "Silhouette"),
            div(class = paste("metric-value", css_sil), val_sil)))
  )

  # в”Ђв”Ђ 1. PCA в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  doc_content[[length(doc_content) + 1]] <- h3("1. PCA (Unsupervised Separation)")

  if (!is.null(res$pca)) {
    pca    <- res$pca
    df_pca <- data.frame(
      PC1    = pca$x[, 1], PC2 = pca$x[, 2],
      Group  = res$raw_data$Class,
      Sample = rownames(res$raw_data))
    ve <- round(summary(pca)$importance[2, 1:2] * 100, 1)

    p_pca <- ggplot(df_pca, aes(x = PC1, y = PC2,
                                 color = Group, text = Sample)) +
      geom_point(size = 3, alpha = 0.8) +
      stat_ellipse(level = 0.95, linetype = "dashed", alpha = 0.4) +
      labs(x = paste0("PC1 (", ve[1], "%)"),
           y = paste0("PC2 (", ve[2], "%)"),
           title = paste0("PCA - ", sig_name)) +
      theme_minimal() +
      scale_color_brewer(palette = "Set1")

    doc_content[[length(doc_content) + 1]] <- ggplotly(p_pca, tooltip = "text")
  } else {
    doc_content[[length(doc_content) + 1]] <- p("PCA unavailable for this signature.")
  }

  # в”Ђв”Ђ 2. Heatmap в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  doc_content[[length(doc_content) + 1]] <- h3("2. Hierarchical Clustering")

  annot_col <- data.frame(Group = res$raw_data$Class)
  rownames(annot_col) <- rownames(res$raw_data)

  p_heat <- tryCatch({
    heatmaply::heatmaply(
      res$beta_matrix,
      limits          = c(0, 1),
      colors          = colorRampPalette(c("navy", "white", "firebrick"))(100),
      col_side_colors = annot_col,
      row_dend_left   = TRUE,
      col_dend_top    = TRUE,
      branches_lwd    = 0.2,
      labRow = if (nrow(res$beta_matrix) > 50) NULL
               else rownames(res$beta_matrix),
      main = paste0("Clustering - ", res$n_cpgs, " CpGs")
    )
  }, error = function(e) NULL)

  if (!is.null(p_heat))
    doc_content[[length(doc_content) + 1]] <- p_heat
  else
    doc_content[[length(doc_content) + 1]] <- p("Heatmap rendering failed.")

  # в”Ђв”Ђ 3. SVM Performance в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  doc_content[[length(doc_content) + 1]] <- h3("3. SVM Performance")

  if (!svm_ok) {
    doc_content[[length(doc_content) + 1]] <- div(
      class = "warn-box",
      strong("SVM analysis not possible."),
      br(),
      "Sample size too small or classes too unbalanced.",
      " Rely on PCA and Heatmap for qualitative assessment."
    )
  } else {
    preds <- res$svm_model$pred %>%
      filter(C == res$svm_model$bestTune$C)
    cm    <- confusionMatrix(preds$pred, preds$obs)
    df_cm <- as.data.frame(cm$table)

    p_cm <- ggplot(df_cm, aes(x = Reference, y = Prediction,
                               fill = Freq)) +
      geom_tile() +
      geom_text(aes(label = Freq), size = 6, color = "white") +
      scale_fill_gradient(low = "#868e96", high = "#0275d8") +
      labs(title = "Confusion Matrix") +
      theme_minimal()

    p_roc <- NULL
    if (res$metric_used == "ROC") {
      roc_obj <- tryCatch(
        pROC::roc(preds$obs, preds[[levels(preds$obs)[1L]]]),
        error = function(e) NULL)

      if (!is.null(roc_obj)) {
        roc_df <- data.frame(
          Specificity = roc_obj$specificities,
          Sensitivity = roc_obj$sensitivities)
        auc_val <- round(roc_obj$auc, 3)

        p_roc <- ggplot(roc_df,
                        aes(x = 1 - Specificity, y = Sensitivity)) +
          geom_line(color = "#d9534f", linewidth = 1) +
          geom_abline(intercept = 0, slope = 1,
                      linetype = "dashed", color = "grey") +
          labs(title = paste0("ROC Curve (AUC = ", auc_val, ")")) +
          theme_minimal()
      }
    }

    if (is.null(p_roc))
      p_roc <- ggplot() +
        annotate("text", x = 0.5, y = 0.5,
                 label = "ROC unavailable (Accuracy mode)") +
        theme_void()

    doc_content[[length(doc_content) + 1]] <- div(
      class = "row",
      div(class = "col-sm-6", ggplotly(p_cm)),
      div(class = "col-sm-6", ggplotly(p_roc))
    )
  }

  # в”Ђв”Ђ 4. Variable Importance в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
  doc_content[[length(doc_content) + 1]] <- h3("4. Top Discriminative CpGs")

  if (!svm_ok) {
    doc_content[[length(doc_content) + 1]] <- p(
      "Requires a valid SVM model. See above.")
  } else {
    imp_obj <- tryCatch(
      varImp(res$svm_model, scale = FALSE),
      error = function(e) NULL)

    if (!is.null(imp_obj) && nrow(imp_obj$importance) > 0) {
      imp <- imp_obj$importance

      imp$Score <- if ("Overall" %in% colnames(imp)) imp$Overall
                   else apply(imp, 1, max)

      imp$RawName <- rownames(imp)
      known <- names(res$genes_map)

      recover_name <- function(x) {
        if (x %in% known) return(x)
        x2 <- gsub("\\.", "-", x)
        if (x2 %in% known) return(x2)
        x3 <- gsub("^X", "", x)
        if (x3 %in% known) return(x3)
        x
      }
      imp$CleanCpG <- vapply(imp$RawName, recover_name, character(1))

      imp$Gene <- res$genes_map[imp$CleanCpG]
      imp$Gene[is.na(imp$Gene) | imp$Gene == ""] <- "Unknown"
      imp$Label <- paste0(imp$CleanCpG, " (", imp$Gene, ")")

      plot_data <- imp %>% arrange(desc(Score)) %>% head(20)
      plot_data$Label <- factor(plot_data$Label,
                                levels = rev(plot_data$Label))

      p_imp <- ggplot(plot_data, aes(x = Label, y = Score)) +
        geom_bar(stat = "identity", fill = "#d9534f", alpha = 0.8) +
        coord_flip() +
        labs(title = "Top 20 Discriminative CpGs",
             x = "", y = "Variable Importance") +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 10))

      doc_content[[length(doc_content) + 1]] <- ggplotly(p_imp)
    } else {
      doc_content[[length(doc_content) + 1]] <- p(
        "Variable importance could not be calculated.")
    }
  }

  doc_content[[length(doc_content) + 1]] <- hr()
}

tagList(doc_content)
```

# Session Info

```{r}
sessionInfo()
```

